Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From Reionization Epoch

### 2014-09-25 foreground_delay_spectrum_modeling.tex
From v0.6.2: ===========
(*) Made minor changes (danny's comments) in the first paragraph of Introduction
(*) Incorporated Judd's suggested changes.
(*) Modified the figures a little bit.
(*) Removed the bandpass figure
(*) Tweaked the title a little bit
(*) Submitted to MWA EoR collaboration

\shorttitle{Foreground signatures in EoR power spectra}
\shortauthors{Thyagarajan et~al.}

@@

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Signatures of Foreground Sky in Power Spectrum of Redshifted Neutral Hydrogen from the Epoch of Reionization}

@@

%% Author list
\author{
%% Lead Authors
Nithyanandan~Thyagarajan\ASU,
Daniel~C.~Jacobs\ASU,
Judd~D.~Bowman\ASU,
%% Supporting authors
% A.~P.~Beardsley\UW,
% J.~Pober\UW,
% I.~S.~Sullivan\UW,
%% Builders' list
G.~Bernardi\SKASA,
F.~Briggs\ANU,
R.~J.~Cappallo\Haystack, 
B.~E.~Corey\Haystack, 
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
R.~Goeke\MIT,
L.~J.~Greenhill\CfA,
B.~J.~Hazelton\UW, 
M.~Johnston-Hollitt\Victoria,
D.~L.~Kaplan\UWisc, 
J.~C.~Kasper\UMichigan$^,$\CfA, 
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack, 
M.~J.~Lynch\Curtin, 
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, 
A.~Roshi\NRAO, 
N.~Udaya~Shankar\RRI, 
K.~S.~Srivani\RRI, 
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~R.~Whitney\Haystack, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT
}

@@

Detection of 21~cm H{\sc i} emission from the Epoch of Reionization (EoR), at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array (MWA) observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact objects dominate the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space.  Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate a large fraction of foreground contamination in redshifted 21~cm power spectrum analysis.

@@

After the CMB epoch, the period in the history of the universe often referred to as the {\it dark ages}, followed the epoch of reionization (EoR). This was a period of non-linear growth of density perturbations and astrophysical evolution. Studying the epoch of reionization holds the key to understanding this evolution. Direct observation of redshifted 21~cm spin flip transition of H{\sc i} has been identified as a useful probe of the EoR \citep{sun72,sco90,mad97,toz00,ili02} and has recently emerged as a very promising experiment to fill the gaps in our understanding of the universe's history.

@@

A key challenge in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components most likely to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\overline{\mathbf{x}}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\eta$ and $k_\parallel$, denoting spatial scales along line of sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, the boundaries of which are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. Spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\eta$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to operating bandwidth. Region of Fourier space excluding the {\it foreground wedge} and the narrow strips is usually called the {\it EoR window}. These are shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis denoted by $|\overline{\mathbf{x}}|$ (antenna spacing in m) or $k_\perp$ (transverse wavenumber in units of $h$~Mpc$^{-1}$) are equivalent and represent transverse spatial scales. Similarly, the $y$--axis denoted by $\eta$ (delay in s or $\mu$s) or equivalently, $k_\parallel$ (line--of--sight wavenumber in units of $h$~Mpc$^{-1}$) represents spatial scales along the line of sight. The dark shaded region is often referred to as {\it foreground wedge} where smooth spectrum foregrounds reside. The boundaries of the {\it foreground wedge} (solid lines) are given by the light travel time for corresponding antenna spacings and are referred to as horizon delay limits. In principle, region outside the wedge is expected to be relatively free of foreground contamination and is called the {\it EoR window} (light and medium shades). In practice, this contamination is also high in the narrow unshaded white regions (bound by solid and dashed lines) due to convolution by instrumental transfer function. The width of this narrow region is determined by inverse of operating bandwidth. Due to an undesirable spectral response of the instrument used in this study, namely the MWA, we identify a narrower region of high significance to EoR studies (medium shade) and hereafter refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

The terms antenna spacing and baseline are synonymous in radio interferometry. The latter term is also used broadly to refer to the antenna pair system in a radio interferometer array. We use these terms interchangeably in this paper.

@@

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30~MHz band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg wide and multiple secondary lobes for the tile. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The bandpass properties used in this paper are illustrated in Figure~\ref{fig:bandpass}. The MWA bandpass of width 30.72~MHz is divided coarsely into 24 1.28~MHz chunks (dashed lines connecting solid circles) with each chunk weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. The bandpass correction ({\it plus} symbols connected by dotted lines) corrects for the shape of the coarse channels. After this correction, the channels at the edges of these chunks have been flagged (solid lines) because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties.eps}
\caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8--tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The bandpass correction factors applied are shown ({\it plus} symbols connected by dashed lines). The solid line shows the flagged (gaps) and gain--corrected bandpass shape. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
\end{figure}

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0h, Dec = -30\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in an MWA tile. These beamformer delay settings can be changed only in discrete steps. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

@@

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at -1.92~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST) in hours. The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA) in degrees. Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. A ``drift and shift'' scheme is used. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0h, Dec = -30\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
\end{figure}

@@

\subsection{Delay Spectrum after Delay Deconvolution}\label{sec:data-delay-spectrum}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two 112~s Murchison Widefield Array observations of the EoR field at RA = 0h, Dec = -30\arcdeg at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. To get the delay spectrum, each baseline is Fourier transformed into delay space ($y$--axis) and the amplitude plotted in order of baseline length ($x$--axis). For reference, the $k$--space equivalent of $\eta$ and $|\overline{\mathbf{x}}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The various features in this region depend on antenna spacing orientation, secondary lobe structure of the power--pattern, and the sky at that instant. \label{fig:fhd_data}}
\end{figure}

@@

We describe power--pattern and foreground models we have used in our simulations. %todo

\subsection{Power--pattern}\label{sec:power_pattern}

The power--pattern of the MWA tile ($A(\overline{\mathbf{l}},f)$ in equation~\ref{eqn:obsvis}) is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
\caption{Models of MWA tile power--pattern in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates for the {\it off--zenith} (top) and {\it zenith} pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above an infinite ground plane. Random fluctuations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown on the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

@@

A low resolution version of the diffuse foreground model used is shown in Figure~\ref{fig:DSM}. Contours of the MWA tile power--pattern shown in Figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power gain is significantly high ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has set and the contour level of power--pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature (in K) of the diffuse foreground model at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities (in Jy) of bright compact objects at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The locations of bright compact objects appear to be random.\label{fig:CSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_sim_delay_spectrum_snapshots.eps}
\caption{Simulated delay spectrum amplitudes $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz) for {\it off--zenith} (top) and {\it zenith} (bottom) pointings. White lines mark the boundaries of {\it foreground wedge} determined by horizon delay limit and antenna spacing. The axes and color scale are identical to those in Figure~\ref{fig:fhd_data}. The features resemble those obtained with MWA data shown in Figure~\ref{fig:fhd_data}. The Galactic center and Galactic plane prominently visible in the {\it off--zenith} pointing makes the emission in the {\it foreground wedge} brighter relative to that in the {\it zenith} pointing. \label{fig:sim_data}}
\end{figure}

>>>>

\shorttitle{Foreground signatures and mitigation in EoR H{\sc i} power spectra}
\shortauthors{Thyagarajan et~al.}

@@

%% \title{Detecting Epoch of Reionization in redshifted 21~cm: contamination in EoR window}
%% \title{Limits on the detection of the Epoch of Reionization from MWA observations of the redshifted 21~cm line}
\title{Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From Reionization Epoch}

@@

%% Author list
\author{
%% Lead Authors
Nithyanandan~Thyagarajan\ASU,
Daniel~C.~Jacobs\ASU,
Judd~D.~Bowman\ASU,
%% Supporting authors
% A.~P.~Beardsley\UW,
% J.~Pober\UW,
% I.~S.~Sullivan\UW,
%% Builders' list
G.~Bernardi\SKASA,
F.~Briggs\ANU,
R.~J.~Cappallo\Haystack, 
B.~E.~Corey\Haystack, 
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
R.~Goeke\MIT,
L.~J.~Greenhill\CfA,
B.~J.~Hazelton\UW, 
M.~Johnston-Hollitt\Victoria,
D.~L.~Kaplan\UWisc, 
J.~C.~Kasper\UMichigan$^,$\CfA, 
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack, 
M.~J.~Lynch\Curtin, 
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, 
A.~Roshi\NRAO, 
N.~Udaya~Shankar\RRI, 
K.~S.~Srivani\RRI, 
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~R.~Whitney\Haystack, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
and~MWA~EoR~members
}

@@

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact objects dominate the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space.  Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate a large fraction of foreground contamination in redshifted 21~cm power spectrum analysis.

@@

At the end of the recombination epoch, the universe turned completely neutral. This period, referred to as the {\it Dark Ages} in the universe's history, is characterized by collapse of matter under the influence of gravity. And it ended with the formation of the first stars and galaxies which started emitting ultra--violet and X--ray radiation, thereby reionizing the neutral medium in their surroundings. This commenced the epoch of reionization (EoR) --- a period of non--linear growth of matter density perturbations and astrophysical evolution. Studying the epoch of reionization holds the key to understanding this evolution.

Observation of redshifted 21~cm spin flip transition of H{\sc i} has been identified as a direct probe of this epoch \citep{sun72,sco90,mad97,toz00,ili02}. Detecting this signal has recently emerged as a very promising experiment to fill the gaps in our understanding of the universe's history.

@@

A key challenge in the statistical detection of power spectrum of this redshifted H{\sc i} 21~cm signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness \citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components most likely to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\overline{\mathbf{x}}|$ and $k_\perp$, denoting spatial scales in the transverse direction, form the $x$--axis. $\eta$ and $k_\parallel$, denoting spatial scales along line of sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge}, whose boundaries are determined by antenna spacing and the light travel time across it. These boundaries are called horizon delay limits and are shown by solid lines. Spectral transfer function of the instrument convolves the {\it foreground wedge} and stretches it further (unshaded narrow strips bounded by solid and dashed lines) along $\eta$--axis. This effect has been quantified in \citet{thy13}. The width of this narrow strip is inversely proportional to operating bandwidth. Region of Fourier space excluding the {\it foreground wedge} and the narrow strips is usually called the {\it EoR window}. These are shown in light and medium shades. However, the instrument used in our study, namely the MWA, has a passband constructed using coarse channels with a characteristic shape giving rise to periodic gratings thereby resulting in repetitions of the {\it foreground wedge} at multiples of 0.78~$\mu$s. The underlying cause of this effect will be discussed later in this paper.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

% The terms antenna spacing and baseline are synonymous in radio interferometry. The latter term is also used broadly to refer to the antenna pair system in a radio interferometer array. We use these terms interchangeably in this paper.

@@

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg wide and multiple secondary lobes for the tile. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The MWA passband of width 30.72~MHz is divided coarsely into 24 1.28~MHz chunks with each chunk weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. After correcting for the shape of these coarse channels, the channels at the edges of these chunks are flagged because they are known to be possibly contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

% The bandpass properties used in this paper are illustrated in Figure~\ref{fig:bandpass}. The MWA bandpass of width 30.72~MHz is divided coarsely into 24 1.28~MHz chunks (dashed lines connecting solid circles) with each chunk weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. The bandpass correction ({\it plus} symbols connected by dotted lines) corrects for the shape of the coarse channels. After this correction, the channels at the edges of these chunks have been flagged (solid lines) because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

% \begin{figure}[htb]
% \centering
% \includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties.eps}
% \caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8--tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The bandpass correction factors applied are shown ({\it plus} symbols connected by dashed lines). The solid line shows the flagged (gaps) and gain--corrected bandpass shape. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
% \end{figure}

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0h, Dec = $-30$\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

@@

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at $-1.92$~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST). The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA). Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0h, Dec = $-30$\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
\end{figure}

@@

\subsection{Deconvolved Delay Spectrum}\label{sec:data-delay-spectrum}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two observations of the EoR field (RA = 0h, Dec = $-30$\arcdeg) at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. To get the delay spectrum, each baseline is Fourier transformed into delay space ($y$--axis) and the amplitude plotted in order of baseline length ($x$--axis). For reference, the $k$--space equivalent of $\eta$ and $|\overline{\mathbf{x}}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. The logarithmic color scale (shown at the right) is common to both panels. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The various features in this region depend on antenna spacing orientation, secondary lobe structure of the power--pattern, and the sky at that instant. \label{fig:fhd_data}}
\end{figure}

@@

We describe the instrumental model and foreground models we use in our simulations.

\subsection{Instrument Model}\label{sec:instrument_model}

The instrument model consists primarily of the tile beam power--pattern, $A(\overline{\mathbf{l}},f)$ (see equation~\ref{eqn:obsvis}). It is modeled as a mutually--coupled 4--by--4 dipole array with the overall power--pattern of each individual dipole calculated via finite element electromagnetic simulation \citep{sut14}. To speed up simulation we find that a phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane provides a very good approximation to the full simulation, hence we use the idealized dipoles. We also assume that each individual dipole signal has random delay fluctuations of rms 0.05~ns, a number in line with the known repeatability and stability level of the analog signal chain \citep{bow07b}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
\caption{Models of MWA tile power--pattern for {\it off--zenith} (top) and {\it zenith} pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above an infinite ground plane. Random fluctuations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown on the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

@@

A low resolution version of the diffuse foreground model used is shown in Figure~\ref{fig:DSM}. Contours of the MWA tile power--pattern shown in Figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the westward sky and the MWA tile's power gain is significant ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has set and the contour level of power--pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature of the diffuse foreground model at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities of bright compact objects at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The locations of bright compact objects appear to be random.\label{fig:CSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_sim_delay_spectrum_snapshots.eps}
\caption{Same as Figure~\ref{fig:fhd_data}, but generated from simulated visibilities using the model described in \S\ref{sec:modeling}. Note the similarity between the broad features to those in Figure~\ref{fig:fhd_data}. \label{fig:sim_data}}
\end{figure}
